Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-09-12 latex/Chapters/Rubiks.tex
spell check

\Section{2x2x2}
All the following experiments on the 2x2x2 \textbf{RC} have been run using my Solver's \textit{performance\_test} (see \ref{sec:codesolvers}): for each level of difficulty, the test present the solvers with the same 100 random configurations (for fairness and to increase the significance of the results). The levels of difficulty (number of scrambles from goal) used here go from 2 to 20 in step of 2. The tests were also run with perfect shuffling (difficulty = $\infty$ on the graphs), corresponding to taking scrambling to $\infty$ as explained in \ref{Theory:222RCSSS}).

\Subsection{BFS}
For \textbf{BFS} I used a step of 1 on the difficuly and stopped at 5 as it was clear it would not go much further. Its run time and expanded nodes are increasing linearly with difficulty, and it was already taking an average close to a minute for 5 shuffles from goal.

@@

\Subsection{Deep Reinforcement Learner \& Deep Q Learner}
\label{sec:S33DRL}
To get my \textbf{DRL} and \textbf{DQL} heuristics, I have run the DeepReinforcementLearner and DeepQLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency}) or when $\frac{MSE}{max target}$ falls below 0.1\% (\textit{update\_target\_network\_threshold=1e-3})
\item Stop at \textit{nb\_epochs}=11,000 or if the max target has not increased more than 1\% (\textit{max\_target\_uptick}) in the last 5,500 epochs (\textit{max\_target\_not\_increasing\_epochs\_pct}=0.5)
\item At each target net update (\textit{training\_data\_every\_epoch}=False), generate \textit{nb\_sequences}=10,000 sequences of puzzles comprised of a 15-scramble puzzle along with its path to goal (\textit{nb\_shuffles}=15).
\item Fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers (\textit{layers\_description}=(600,300,100)), an RMSProp optimiser (\textit{optimiser=rms\_prop}), \textit{scheduler=exponential\_scheduler} with \textit{gamma\_scheduler}=0.9999 and \textit{learning\_rate}=0.1\%. The networks architecture are shown in figure \ref{fig:222RCnets}.
\end{itemize}

@@

\noindent Graphs \ref{fig:222RCDRL} and \ref{fig:222RCDQL} show both learners' following metrics tracked over the epochs:
\begin{itemize}
\item learning rate: decreasing due to the scheduler and reset at each network update.
\item max target: an interesting dynamic during learning is that since targets are constructed using value-iteration update, starting from an all 0 network, the max target increases by at most about 1 at every network update, until we reach God's number. The first network only learns to differentiate goal from non-goal, the second one learns how to differentiate goals, cost 1 and cost 2+ \dots. This is why I introduced \textit{update\_target\_network\_threshold} discussed above, to makes sure we do not spend time initially training the left-hand-side network when it is easy. \textit{max\_target\_uptick} and \textit{max\_target\_not\_increasing\_epochs\_pct} make sure that we stop training when the max target does not grow anymore.
\item loss: MSE for \textbf{DRL}, MSE + CrossEntropyLoss for \textbf{DQL}
\item loss divided by max target: the exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 1\% means little if the we do not compare it to the possible range of the cost-to-go.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the learner. This is an interesting quantity, as it tells us how well the solvers are able to generalise. From both graphs (\ref{fig:222RCDRL} \& \ref{fig:222RCDQL}) we can see that the learners saw less than 1\% of all puzzles. It is good to keep in mind however that \textit{seen} does not mean really much here, since the learning is \textit{unsupervised}. This is quite in contrast to the DeepLearner, whose traiing data consist of puzzles with their cost-to-go computed by a \textit{teacher} (optimal or other efficient solver).
\end{itemize}

@@

I did not get to run very many experiments on the 3x3x3 \textbf{RC}. The results of the performance test (50 random configurations for levels of difficulty from 2 to 40 by increment of 2 as well as $\infty$ shuffling) can be found on figure \ref{fig:333RCPerformance}.
\begin{itemize}
\item Here again, Kociemba does not show optimality, since we can see that my A$^{*}[Kociemba]$, that is A$^{*}$ using Kociemba as a heuristic, performs much better than Kociemba (solutions about 20\% shorter at $\infty$ shuffling). This is not surprising though since Kociemba 3x3x3 does not claim to be optimal and is meant to find solution in the 40 moves at most (which is what I have observed indeed). A$^{*}[Kociemba]$ never took more than 30 moves on any of the cubes presented to it, which is not bad given the known God number of 26 in the quarter-turn-metric (as discussed earlier in \ref{Theory:333RCSSS})
\item A$^{*}$[\textbf{DL}] managed to solve all puzzles scrambled less than 10 times, and found shorter worst case solutions than even A$^{*}[Kociemba]$ on which it was trained. Within the 2 hours time out, it failed to solve one of the 50 cubes for difficulty 10, and two of the 50 cubes of difficulty 12, so i did not try further than that.
\item A$^{*}$[\textbf{DRL}] only managed to solve all puzzles of difficulty less than 6 but at least did find only optimal solutions (compared with \textbf{BFS}).
\end{itemize}

>>>>

\Section{2x2x2}
All the following experiments on the 2x2x2 \textbf{RC} have been run using my Solver's \textit{performance\_test} (see \ref{sec:codesolvers}): for each level of difficulty, the test presents the solvers with the same 100 random configurations (for fairness and to increase the significance of the results). The levels of difficulty (number of scrambles from goal) used here go from 2 to 20 in step of 2. The tests were also run with perfect shuffling (difficulty = $\infty$ on the graphs), corresponding to taking scrambling to $\infty$ as explained in \ref{Theory:222RCSSS}).

\Subsection{BFS}
For \textbf{BFS} I used a step of 1 on the difficulty and stopped at 5 as it was clear it would not go much further. Its run time and expanded nodes are increasing linearly with difficulty, and it was already taking an average close to a minute for 5 shuffles from goal.

@@

\Subsection{Deep Reinforcement Learner \& Deep Q Learner}
\label{sec:S33DRL}
To get my \textbf{DRL} and \textbf{DQL} heuristics, I have run the DeepReinforcementLearner and DeepQLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency}) or when $\frac{MSE}{max target}$ falls below 0.1\% (\textit{update\_target\_network\_threshold=1e-3})
\item Stop at \textit{nb\_epochs}=11,000 or if the max target has not increased more than 1\% (\textit{max\_target\_uptick}) in the last 5,500 epochs (\textit{max\_target\_not\_increasing\_epochs\_pct}=0.5)
\item At each target net update (\textit{training\_data\_every\_epoch}=False), generate \textit{nb\_sequences}=10,000 sequences of puzzles comprised of a 15-scramble puzzle along with its path to goal (\textit{nb\_shuffles}=15).
\item Fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers (\textit{layers\_description}=(600,300,100)), an RMSProp optimiser (\textit{optimiser=rms\_prop}), \textit{scheduler=exponential\_scheduler} with \textit{gamma\_scheduler}=0.9999 and \textit{learning\_rate}=0.1\%. The networks architectures are shown in figure \ref{fig:222RCnets}.
\end{itemize}

@@

\noindent Graphs \ref{fig:222RCDRL} and \ref{fig:222RCDQL} show both learners' following metrics tracked over the epochs:
\begin{itemize}
\item learning rate: decreasing due to the scheduler and reset at each network update.
\item max target: an interesting dynamic during learning is that since targets are constructed using value-iteration update, starting from an all 0 network, the max target increases by at most about 1 at every network update, until we reach God's number. The first network only learns to differentiate goal from non-goal, the second one learns how to differentiate goals, cost 1 and cost 2+ \dots. This is why I introduced \textit{update\_target\_network\_threshold} discussed above, to makes sure we do not spend time initially training the left-hand-side network when it is easy. \textit{max\_target\_uptick} and \textit{max\_target\_not\_increasing\_epochs\_pct} make sure that we stop training when the max target does not grow anymore.
\item loss: MSE for \textbf{DRL}, MSE + CrossEntropyLoss for \textbf{DQL}
\item loss divided by max target: the exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 1\% means little if the we do not compare it to the possible range of the cost-to-go.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the learner. This is an interesting quantity, as it tells us how well the solvers are able to generalise. From both graphs (\ref{fig:222RCDRL} \& \ref{fig:222RCDQL}) we can see that the learners saw less than 1\% of all puzzles. It is good to keep in mind however that \textit{seen} does not mean really much here since the learning is \textit{unsupervised}. This is quite in contrast to the DeepLearner, whose training data consist of puzzles with their cost-to-go computed by a \textit{teacher} (optimal or other efficient solver).
\end{itemize}

@@

I did not get to run very many experiments on the 3x3x3 \textbf{RC}. The results of the performance test (50 random configurations for levels of difficulty from 2 to 40 by increment of 2 as well as $\infty$ shuffling) can be found on figure \ref{fig:333RCPerformance}.
\begin{itemize}
\item Here again, Kociemba does not show optimality, since we can see that my A$^{*}[Kociemba]$, that is A$^{*}$ using Kociemba as a heuristic, performs much better than Kociemba (solutions about 20\% shorter at $\infty$ shuffling). This is not surprising though since Kociemba 3x3x3 does not claim to be optimal and is meant to find solution in the 40 moves at most (which is what I have observed indeed). A$^{*}[Kociemba]$ never took more than 30 moves on any of the cubes presented to it, which is not bad given the known God number of 26 in the quarter-turn-metric (as discussed earlier in \ref{Theory:333RCSSS})
\item A$^{*}$[\textbf{DL}] managed to solve all puzzles scrambled less than 10 times, and found shorter worst-case solutions than even A$^{*}[Kociemba]$ on which it was trained. Within the 2 hours time-out, it failed to solve one of the 50 cubes for difficulty 10, and two of the 50 cubes of difficulty 12, so i did not try further than that.
\item A$^{*}$[\textbf{DRL}] only managed to solve all puzzles of difficulty less than 6 but at least did find only optimal solutions (compared with \textbf{BFS}).
\end{itemize}

### 2022-09-12 latex/Chapters/SlidingPuzzle.tex
spell check

As mentioned in chapter \ref{sec:Puzzles}, the state space cardinality for the \textbf{SP} grows quickly with n and m. The only dimensions which have less than 239.5 millions states are shown in table \ref{tab:smallSP}.

\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 12 & 360 & 20,160 & 1,814,400 \\
3              &   &   & 181,440 &  &    \\
\end{tabular}
\caption{\label{tab:smallSP}\# puzzles for \textit{small} dimensions}
\end{center}
\end{table}
\noindent Using my PerfectLearner with A* and Manhattan heuristic (see \ref{PLSS} for example of how to run this), I obtained the following God numbers (table \ref{tab:smallSPGN}) for these dimensions:
\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 6 & 21 & 36 & 55* \\
3              &   &   & 31 &  &    \\
* \tiny{provisional result}
\end{tabular}
\caption{\label{tab:smallSPGN}God numbers for \textit{small} dimensions}
\end{center}
\end{table}
\noindent Among those dimensions, 2x5 has the largest state space and my result for it is still provisional, as I have only solved about 1.2 out of 1.8 million puzzles at the time of writing, despite running a combined 600 hours (mostly over 16 cores, except for a full week-end run on a 72-core c5.18xlarge Amazon EC2 instance). It is well possible that I haven't yet bumped into the hardest 2x5 configuration! The perfectLearner also keeps track of the hardest puzzle it has encountered (i.e. a configuration whose optimal solution has a cost equal to the God number). I obtained the following hardest puzzles for each of the 5 \textit{small} dimensions:
\\
\begin{three}
\centering
\setrow{2}{,3}
\setrow{1}{2,1}
\end{three}
\begin{five}
\setrow{2}{4,5,}
\setrow{1}{1,2,3}
\end{five}
\begin{seven}
\setrow{2}{,7,2,1}
\setrow{1}{4,3,6,5}
\end{seven}
\begin{nine}
\setrow{2}{,9,3,7,1}
\setrow{1}{5,4,8,2,6}
\end{nine}
\begin{eight}
\setrow{3}{8,6,7}
\setrow{2}{2,5,4}
\setrow{1}{3,,1}
\end{eight}

@@

\Subsection{Perfect learner}
\label{sec:S33PL}
As discussed in the previous section section \ref{sec:SPLowDimension}, the 3x3 \textbf{SP} is one of the cases I have been able to solve perfectly, since it only has 181,440 possible configurations. Its God number is only 31, which definitely makes it manageable. However, this is already an intermediary size, large enough that it is worth trying and comparing a few different methods, including \textbf{DL}, \textbf{DRL} and \textbf{DQL}. To start with, I ran the PerfectLearner and the results are shown below in figure \ref{fig:33SPPerfectLearning}. It is interesting to note that there are only two hardest configurations (of cost 31) and 221 configurations of cost 30.

@@

\Subsection{MCTS DQL}
\label{sec:MTCSHyperParamsEffect}
In this subsection I explore the effect on \textbf{MCTS} of the trade-off hyper-parameter $c$ as well as of trimming the resulting tree via \textbf{BFS}. (see the threory section \ref{sec:TheoryMCTS} as well as the implementation section \ref{sec:codesolvers} for more details). To that effect, I have run my Solver performance test with four different values of $c$ (0, 1, 10 and 100), and in each case, with and without \textbf{BFS} trimming. A$^{*}[Manhattan++]$ is also presented to have an optimal solver benchmark. For each level of difficulty, the test present the solvers with the same 150 random configurations (for fairness and to increase the significance of the results). The levels of difficulty (number of scrambles from goal) used here go from 5 to 50 in step of 5. The tests were also run with perfect shuffling (difficulty = $\infty$ on the graphs), corresponding to taking scrambling to $\infty$ as explained in \ref{Theory:SPSSS}). I used a \textit{time\_out} of 5 minutes, which none of the solvers hit on any of the (1,650) configurations they were presented with.
The detailed results (percentage of configurations solved, optimality score, median and max cost, run time and expanded nodes) are shown on the next page \ref{fig:33SPPerformanceMCTS}, but let me make here some remarks on those results:
\begin{itemize}
\item $c=0$ makes no exploration as it simply follows the path of lowest cost-to-go. As $c$ increases, we expect more exploration, and therefore higher run time, more expanded nodes, and higher quality solutions. This is what we observe, where $c=0$ converges (for difficulty=$\infty$) to solutions on average three times longer than the optimal solver but take not much longer to obtain. At the other extreme, $c=100$ takes almost 2 orders of magnitude more time to obtain a solution, but these solutions are only 20-25\% longer than that of the optimal solver
\item trimming via \textbf{BFS} takes hardly any time, and improves solutions' lengths by about 10\% at high values of $c$. At low value of $c$ there is little benefit as expected (since the trees are deep but not bushy at all and have much fewer leaves and branches).
\item The quality of solutions is not amazing, but then given the algorithm has not really any mechanism to guarantee optimality this is not very surprising. 
\item Notice that the number of expanded nodes can be compared within \textbf{MCTS} but given the algorithm is totally different to A$^{*}$ it makes little sense to compare between the different families of solvers. Also one point to note is that \textbf{MCTS} would be fairly easy to multithread (within the same process) using an appropriate language such as C++, and this is what \cite{https://doi.org/10.48550/arxiv.1805.07470} have done, and probably why they get considerably faster and better results.
\item Despite the name \textit{Monte Carlo}, there is no randomness (or even peudo-randomness) in this algorithm, since the paths generated at each iteration are really just a deterministic function of the output of the \textbf{DQL} network (or QHeuristic in my code-jargon).
\end{itemize}

@@

\begin{itemize}
\item As expected, \textbf{BFS} becomes quickly hopeless.
\item The naive solver is obvioulsy by far the fastest but finds poor solutions (about 2.5 longer on average than optimal).
\item As discussed in the previous section, \textbf{MCTS} is slow, and even with much exploration and \textbf{BFS} tree trimming found solution about 25\% longer than optimal.
\item I originally was naively hoping a pure convolutional network would do well but did not manage to get the loss to converge. I later realized that this is quite different from the typical image patterns recognition as there is less direct translation and scale invariants at play here, and decided instead to try combining convolutional with fully connected layers in parallel in order to appropriately value 2x2 patterns together with their position on the sliding-puzzle. That did improve things, but overall I saw little difference in performance from my simpler 3-hidden-layer fully connected network.
\item \textbf{DRL} and \textbf{DQL} heuristics, though not as efficient in terms node-expansion as the perfect and the \textbf{DL} heuristics, were rather surprisingly doing better than Manhattan++. They ended up \textit{seeing} around 25\% of all the possible configurations during training (in an \textit{unsupervised} way though) so did show some reasonable amount of out-of-sample generalization. Their optimality score was an impressive 99\%+ at all levels of difficuly for \textbf{DRL} and 98\%+ for \textbf{DQL}. Even more interestingly, the \textbf{DQL} heuristic expanded almost 4 times less nodes than \textbf{DRL}, suggesting some benefit from the joint learning of the policy and value function.
\item The \textbf{DL} heuristics (fully connected and convolutional) performed almost on par with the perfect heuristic (optimality score of 99\%+ at all levels of difficulty and nodes expansion only increased by about 12\%), despite seeing only 8\% of the configurations by the end of their training.
\end{itemize}

@@

\begin{center}
\begin{tabular}{l*{5}{c}r}
\hline
\textbf{solver}      & & \textbf{cost} & \textbf{\# expanded nodes} & \textbf{run time (ms)} \\
\hline
AStarSolver[Manhattan]   &   &      31  & 58,859  & 11,327  \\
\hline
AStarSolver[Manhattan++]   &   &      31  & 34,224  & 7,080  \\
\hline
AStarSolver[PerfectHeuristic]  &   & 31 & 1,585 & 202 \\
\hline
AStarSolver[DRLHeuristic]  &   & 31 & 101 & 58 \\
\hline
MCTSSolver[DQLHeuristic][c=0]  &   & 101 & 103 & 456 \\
\hline
MCTSSolver[DQLHeuristic][c=69]  &   & 35 & 2,244 & 8,873 \\
\hline
BFS  &   & - & - & time out \\
\hline
NaiveSlidingSolver  &   & 61 & n/a & 18 \\
\hline
\end{tabular}
\end{center}
On this specific configuration, \textbf{BFS} was unable to complete before the time-out of sixty seconds. In an implementation without duplicate pruning, \textbf{BFS} would time out no matter what the bound is set. Indeed, it would need to explore in the order of $3^{31}$ - roughly 617 trillions - nodes to reach the goal! Even with my implementation which does pruning, it would need to pretty much traverse the entire 3x3 \textbf{SP} tree.
\\
Rather interestingly, the \textbf{DRL} heuristic performs much better than the Manhattan heuristic (not super suprising), but also outperforms the perfect heuristic quite significantly both in terms of run time and of nodes expansion. Obviously there is no guarantee that the perfect heuristic will not be outperformed on some random configuration, and it does on this occasion. However, as we have seen in the previous subection \ref{ssec:33SPSC}, it is not the case on average.
\\
\textbf{MCTS} with $c$ = 0 performs rather poorly, finding the longest solution (even than my Naive solver). It does behave a bit like \textbf{DFS} when $c$ is very small, expect the direction of travel is a bit more informed. I increased $c$, and for values over 69, it always gave me a solution of cost 35, which is not bad at all.
\\
Finally, the naive solver outperforms every other solver in terms of run time, but finds a rather poor solution of 61 moves.

@@

For the 3x4 \textbf{SP}, I ran the performance test (100 random puzzles for difficulty going from 1 to 45 as well as $\infty$ shuffling) on the Naive solver and A$^{*}$ with Manhattan, Manhattan++, \textbf{DRL}, \textbf{DQL} and \textbf{DL} heuristics. Given the 239,500,800 total possible configurations there was no chance for me to compute the perfect heuristic as I did with smaller dimensions, so \textbf{DL} got trained on randomly generated configurations solved by A$^{*}$[Manhattan++]. The network architecture for all \textbf{DxL} was the fully-connected (left-hand-side of figure \ref{fig:33SPnets}), mutatis-mutandis for the input size, which in the general case is $(n.m)^{2}$ when using one-hot-encoding (I always did after a few unsuccessful attempts without it). Detailed results can be found at \ref{fig:34SPPerformance}. Key take-aways are that all \textbf{DxL} outperformed A$^{*}$[Manhattan++] in terms of node expansion and got optimality scores of 99\%+ for all levels of difficulty, despite only seeing 0.009\% of all configurations for \textbf{DL}, and 0.27\% for \textbf{DRL} and \textbf{DQL}. Again this shows very impressive generalisation and performance from the\textbf{DxL} methods.

@@

In the 4x4 case, due to time constraints, I ran the performance test with 100 random puzzles for difficulty going from 5 to 60 except for $\infty$ shuffling where I only ran a handful of configurations (5). I ran the Naive solver as well as A$^{*}$ with Manhattan++, \textbf{DRL} and \textbf{DL} heuristics, again using the same fully-connected network architecture (left-hand-side of figure \ref{fig:33SPnets}). The 4x4 \textbf{SP} has 10,461,394,944,000 total possible configurations so it does start being quite a large state space! Again \textbf{DL} got trained on randomly generated configurations solved by A$^{*}$[Manhattan++]. \textbf{DL} obtained an optimality scores of 97\%+ across all levels of difficulty and \textbf{DRL} managed 99\%+. \textbf{DL} and \textbf{DRL} saw of the order of respectively $10^{-7}$\% and $10^{-5}$\% of all configurations during training, so again showed impressive performance and generalization!

>>>>

As mentioned in chapter \ref{sec:Puzzles}, the state space cardinality for the \textbf{SP} grows quickly with n and m. The only dimensions which have less than 239.5 million states are shown in table \ref{tab:smallSP}.

\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 12 & 360 & 20,160 & 1,814,400 \\
3              &   &   & 181,440 &  &    \\
\end{tabular}
\caption{\label{tab:smallSP}\# puzzles for \textit{small} dimensions}
\end{center}
\end{table}
\noindent Using my PerfectLearner with A* and Manhattan heuristic (see \ref{PLSS} for example of how to run this), I obtained the following God numbers (table \ref{tab:smallSPGN}) for these dimensions:
\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 6 & 21 & 36 & 55* \\
3              &   &   & 31 &  &    \\
* \tiny{provisional result}
\end{tabular}
\caption{\label{tab:smallSPGN}God numbers for \textit{small} dimensions}
\end{center}
\end{table}
\noindent Among those dimensions, 2x5 has the largest state space and my result for it is still provisional, as I have only solved about 1.2 out of 1.8 million puzzles at the time of writing, despite running a combined 600 hours (mostly over 16 cores, except for a full weekend run on a 72-core c5.18xlarge Amazon EC2 instance). It is well possible that I haven't yet bumped into the hardest 2x5 configuration! The perfectLearner also keeps track of the hardest puzzle it has encountered (i.e. a configuration whose optimal solution has a cost equal to the God number). I obtained the following hardest puzzles for each of the 5 \textit{small} dimensions:
\\
\begin{three}
\centering
\setrow{2}{,3}
\setrow{1}{2,1}
\end{three}
\begin{five}
\setrow{2}{4,5,}
\setrow{1}{1,2,3}
\end{five}
\begin{seven}
\setrow{2}{,7,2,1}
\setrow{1}{4,3,6,5}
\end{seven}
\begin{nine}
\setrow{2}{,9,3,7,1}
\setrow{1}{5,4,8,2,6}
\end{nine}
\begin{eight}
\setrow{3}{8,6,7}
\setrow{2}{2,5,4}
\setrow{1}{3,,1}
\end{eight}

@@

\Subsection{Perfect learner}
\label{sec:S33PL}
As discussed in the previous section \ref{sec:SPLowDimension}, the 3x3 \textbf{SP} is one of the cases I have been able to solve perfectly, since it only has 181,440 possible configurations. Its God number is only 31, which makes it manageable. However, this is already an intermediary size, large enough that it is worth trying and comparing a few different methods, including \textbf{DL}, \textbf{DRL} and \textbf{DQL}. To start with, I ran the PerfectLearner and the results are shown below in figure \ref{fig:33SPPerfectLearning}. It is interesting to note that there are only two hardest configurations (of cost 31) and 221 configurations of cost 30.

@@

\Subsection{MCTS DQL}
\label{sec:MTCSHyperParamsEffect}
In this subsection I explore the effect on \textbf{MCTS} of the trade-off hyper-parameter $c$ as well as of trimming the resulting tree via \textbf{BFS}. (see the theory section \ref{sec:TheoryMCTS} as well as the implementation section \ref{sec:codesolvers} for more details). To that effect, I have run my Solver performance test with four different values of $c$ (0, 1, 10 and 100), and in each case, with and without \textbf{BFS} trimming. A$^{*}[Manhattan++]$ is also presented to have an optimal solver benchmark. For each level of difficulty, the test presents the solvers with the same 150 random configurations (for fairness and to increase the significance of the results). The levels of difficulty (number of scrambles from goal) used here go from 5 to 50 in step of 5. The tests were also run with perfect shuffling (difficulty = $\infty$ on the graphs), corresponding to taking scrambling to $\infty$ as explained in \ref{Theory:SPSSS}). I used a \textit{time\_out} of 5 minutes, which none of the solvers hit on any of the (1,650) configurations they were presented with.
The detailed results (percentage of configurations solved, optimality score, median and max cost, run time and expanded nodes) are shown on the next page \ref{fig:33SPPerformanceMCTS}, but let me make here some remarks on those results:
\begin{itemize}
\item $c=0$ makes no exploration as it simply follows the path of lowest cost-to-go. As $c$ increases, we expect more exploration, and therefore higher run time, more expanded nodes, and higher quality solutions. This is what we observe, where $c=0$ converges (for difficulty=$\infty$) to solutions on average three times longer than the optimal solver but take not much longer to obtain. At the other extreme, $c=100$ takes almost 2 orders of magnitude more time to obtain a solution, but these solutions are only 20-25\% longer than that of the optimal solver
\item trimming via \textbf{BFS} takes hardly any time, and improves solutions' lengths by about 10\% at high values of $c$. At low value of $c$ there is little benefit as expected (since the trees are deep but not bushy at all and have much fewer leaves and branches).
\item The quality of solutions is not amazing, but then given the algorithm has not really any mechanism to guarantee optimality this is not very surprising. 
\item Notice that the number of expanded nodes can be compared within \textbf{MCTS} but given the algorithm is totally different to A$^{*}$ it makes little sense to compare between the different families of solvers. Also one point to note is that \textbf{MCTS} would be fairly easy to multithread (within the same process) using an appropriate language such as C++, and this is what \cite{https://doi.org/10.48550/arxiv.1805.07470} have done, and probably why they get considerably faster and better results.
\item Despite the name \textit{Monte Carlo}, there is no randomness (or even peudo-randomness) in this algorithm, since the paths generated at each iteration are really just a deterministic function of the output of the \textbf{DQL} network (or QHeuristic in my code-jargon).
\end{itemize}

@@

\begin{itemize}
\item As expected, \textbf{BFS} becomes quickly hopeless.
\item The naive solver is obviously by far the fastest but finds poor solutions (about 2.5 longer on average than optimal).
\item As discussed in the previous section, \textbf{MCTS} is slow, and even with much exploration and \textbf{BFS} tree trimming found solution about 25\% longer than optimal.
\item I originally was naively hoping a pure convolutional network would do well but did not manage to get the loss to converge. I later realized that this is quite different from the typical image patterns recognition as there is less direct translation and scale invariants at play here, and decided instead to try combining convolutional with fully connected layers in parallel in order to appropriately value 2x2 patterns together with their position on the sliding-puzzle. That did improve things, but overall I saw little difference in performance from my simpler 3-hidden-layer fully connected network.
\item \textbf{DRL} and \textbf{DQL} heuristics, though not as efficient in terms node-expansion as the perfect and the \textbf{DL} heuristics, were rather surprisingly doing better than Manhattan++. They ended up \textit{seeing} around 25\% of all the possible configurations during training (in an \textit{unsupervised} way though) so did show some reasonable amount of out-of-sample generalization. Their optimality score was an impressive 99\%+ at all levels of difficulty for \textbf{DRL} and 98\%+ for \textbf{DQL}. Even more interestingly, the \textbf{DQL} heuristic expanded almost 4 times less nodes than \textbf{DRL}, suggesting some benefit from the joint learning of the policy and value function.
\item The \textbf{DL} heuristics (fully connected and convolutional) performed almost on par with the perfect heuristic (optimality score of 99\%+ at all levels of difficulty and nodes expansion only increased by about 12\%), despite seeing only 8\% of the configurations by the end of their training.
\end{itemize}

@@

\begin{center}
\begin{tabular}{l*{5}{c}r}
\hline
\textbf{solver}      & & \textbf{cost} & \textbf{\# expanded nodes} & \textbf{run time (ms)} \\
\hline
AStarSolver[Manhattan]   &   &      31  & 58,859  & 11,327  \\
\hline
AStarSolver[Manhattan++]   &   &      31  & 34,224  & 7,080  \\
\hline
AStarSolver[PerfectHeuristic]  &   & 31 & 1,585 & 202 \\
\hline
AStarSolver[DRLHeuristic]  &   & 31 & 101 & 58 \\
\hline
MCTSSolver[DQLHeuristic][c=0]  &   & 101 & 103 & 456 \\
\hline
MCTSSolver[DQLHeuristic][c=69]  &   & 35 & 2,244 & 8,873 \\
\hline
BFS  &   & - & - & time out \\
\hline
NaiveSlidingSolver  &   & 61 & n/a & 18 \\
\hline
\end{tabular}
\end{center}
On this specific configuration, \textbf{BFS} was unable to complete before the time-out of sixty seconds. In an implementation without duplicate pruning, \textbf{BFS} would time out no matter what the bound is set. Indeed, it would need to explore in the order of $3^{31}$, roughly 617 trillions nodes to reach the goal! Even with my implementation which does pruning, it would need to pretty much traverse the entire 3x3 \textbf{SP} tree.
\\
Rather interestingly, the \textbf{DRL} heuristic performs much better than the Manhattan heuristic (not super surprising), but also outperforms the perfect heuristic quite significantly both in terms of run time and of nodes expansion. Obviously, there is no guarantee that the perfect heuristic will not be outperformed on some random configuration, and it does on this occasion. However, as we have seen in the previous subsection \ref{ssec:33SPSC}, it is not the case on average.
\\
\textbf{MCTS} with $c$ = 0 performs rather poorly, finding the longest solution (even than my Naive solver). It does behave a bit like \textbf{DFS} when $c$ is very small, expect the direction of travel is a bit more informed. I increased $c$, and for values over 69, it always gave me a solution of cost 35, which is not bad at all.
\\
Finally, the naive solver outperforms every other solver in terms of run time, but finds a rather poor solution of 61 moves.

@@

For the 3x4 \textbf{SP}, I ran the performance test (100 random puzzles for difficulty going from 1 to 45 as well as $\infty$ shuffling) on the Naive solver and A$^{*}$ with Manhattan, Manhattan++, \textbf{DRL}, \textbf{DQL} and \textbf{DL} heuristics. Given the 239,500,800 total possible configurations there was no chance for me to compute the perfect heuristic as I did with smaller dimensions, so \textbf{DL} got trained on randomly generated configurations solved by A$^{*}$[Manhattan++]. The network architecture for all \textbf{DxL} was the fully-connected (left-hand-side of figure \ref{fig:33SPnets}), mutatis-mutandis for the input size, which in the general case is $(n.m)^{2}$ when using one-hot-encoding (I always did after a few unsuccessful attempts without it). Detailed results can be found at \ref{fig:34SPPerformance}. Key take-aways are that all \textbf{DxL} outperformed A$^{*}$[Manhattan++] in terms of node expansion and got optimality scores of 99\%+ for all levels of difficulty, despite only seeing 0.009\% of all configurations for \textbf{DL}, and 0.27\% for \textbf{DRL} and \textbf{DQL}. Again this shows very impressive generalisation and performance from the \textbf{DxL} methods.

@@

In the 4x4 case, due to time constraints, I ran the performance test with 100 random puzzles for difficulty going from 5 to 60 except for $\infty$ shuffling where I only ran a handful of configurations (5). I ran the Naive solver as well as A$^{*}$ with Manhattan++, \textbf{DRL} and \textbf{DL} heuristics, again using the same fully-connected network architecture (left-hand-side of figure \ref{fig:33SPnets}). The 4x4 \textbf{SP} has 10,461,394,944,000 total possible configurations so it does start being quite a large state space! Again \textbf{DL} got trained on randomly generated configurations solved by A$^{*}$[Manhattan++]. \textbf{DL} obtained an optimality score of 97\%+ across all levels of difficulty and \textbf{DRL} managed 99\%+. \textbf{DL} and \textbf{DRL} saw of the order of respectively $10^{-7}$\% and $10^{-5}$\% of all configurations during training, so again showed impressive performance and generalization!
